Wolbachia, Toxoplasma and O. Cordyceps: Extreme Host-Manipulating Microbes That Are Changing the World

### 2015-12-17 paper.tex
Added what is needed for why or why not?
removed the extra refferences section
Titles
added old_leaf refference
EVERYTHING

The bacteria is found in the ovaries and testes of a wide range of arthropods.\cite{Wbio} It is passed on via reproduction, and can only be spread through females. \textit{Wolbachia} are considered mutualistic endosymbionts, and are "required for survival of their hosts."\cite{Wdisc_nem} An interesting thing to note is how universal \textit{Wolbachia}seems to be. It has been demonstrated that \textit{Wolbachia} infects 25-70\% of species of insects.\cite{Wdisc_nem}

To give a bit of background,\textit{Wolbachia} was first discovered by Samuel Burt Wolbach and described by Marshall Hertig in 1924.\cite{Winit} Wolbach had been studying the cause of Rocky Mountain Spotted Fever and by 1916 had shown that \textit{Dermacentor andersoni}, a tick native to the Rocky Mountain area, was the transmitter of the disease.\cite{wolbachia} He was unable to grow the actual bacteria causing the disease, a species of \textit{Rickettsia}, in a cell-free culture leading "him to speculate as to the relationship between the cells of the host and the intracellular parasites."\cite{wolbachia} This is important to note, because the discovery of \textit{Wolbachia} is closely linked to this idea.

@@

Also during the 1960s and '70s scientists noted "unusual structures in the oocytes or in the hypodermis" of filarial nematodes but did not associate them with bacteria.\cite{wolbachia} It wasn't until 1995 that these "unusual structures" were determined to be \textit{Wolbachia}.\cite{Wstruct}

@@

%##### How does it function?
The most prominent of these reproductive changes is cytoplasmic incompatibility which is a "reproductive incompatibility between sperm and egg."\cite{Wbio} \textit{Wolbachia} can only be passed on through eggs, and not sperm, so in order to get around this limitation it has evolved to not allow inter-breeding between members of the same species where only one of them is infected with \textit{Wolbachia}. However, it gets even more complicated than this.

@@

The exact mechanisms for the cause of cytoplasmic incompatibility are not known. What is known is that the \textit{Wolbachia} in the male modify the sperm in such a way that only if the same strain of \textit{Wolbachia} is in the female species to "recover" the sperm will correct fertilization occur.\cite{Wbio}

In many ways \textit{Wolbachia} acts like an encryption algorithm, "encrypting" the sperm sent to the female, and only if the female has the correct decryption algorithm can she get pregnant. This is also why different strains of \textit{Wolbachia} are incompatible, using a wrong decryption algorithm will generate nonsense.

@@

Perhaps the most famous case of parthogenisis caused by \textit{Wolbachia} is in the \textit{Trichogramma} wasps, where all members of the species are female. Weirdly, by treating the wasps with antibiotics to kill of the \textit{Wolbachia} cause "some female parthenogenetic strains ... to revert to production of male progeny."\cite{Wpar_removal} It would appear that \textit{Wolbachia} is merely surpressing the need for male's in certain species. Even stranger is the fact that "phylogenetic evidence suggests that [parthogenisis] has evolved several times independently in these bacteria" which, as one scientist puts it, suggests "a simple biochemical mechanism."\cite{Wbio}

@@

\section*{Toxoplasma}

@@

\textit{Ophiocordyceps unilateralis} was first discovered in 1859 by the British naturalist Alfred Russel Wallace who, probably not coincidently, is best known for independently conceiving the theory of evolution through natural selection. Darwin published his now famous "On the Origin of Species" in the same year, which gives some good context as to where the scientific paradigms were at the time.\cite{darwin}

\section*{Impacts}

@@

- Toxoplasma has been correlated to a rise in schizophernia

The presence of \textit{Wolbachia} in filarial nematodes has raised interesting questions about how \textit{Wolbachia} evolves in its host species. Some scientists discovered that \textit{Wolbachia} "have possibly been lost during evolution along some lineages of filarial nematodes."\cite{Wevolution_loss} The reason this discovery is bizare is because \textit{Wolbachia}'s ability to evolve into the life-cycles of so many different species makes sense if \textit{Wolbachia} gives the infected a fitness advantage over uninfected members. However, if species evolve to get rid of \textit{Wolbachia} then this points to the fact that \textit{Wolbachia} does not supply them with a fitness advantage.

@@

Using \textit{Wolbachia} as an "antidote" for different diseases is novel and has potential to change the world dramatically. The best part is that it is a one-time use antidote, because \textit{Wolbachia} has the tools to infect an entire species through vertical transmission. It may also cause scientists to look at other symbiotic relationships between organisms, and see if they have any potential to help humans. There might be a new era of using living organisms as tools to cure diseases and help humans in other ways.

@@

\section*{References}

>>>>

The bacteria is found in the ovaries and testes of a wide range of arthropods.\cite{Wbio} It is passed on via reproduction, and can only be spread through females. \textit{Wolbachia} are considered mutualistic endosymbionts, and are ``required for survival of their hosts."\cite{Wdisc_nem} An interesting thing to note is how universal \textit{Wolbachia}seems to be. It has been demonstrated that \textit{Wolbachia} infects 25-70\% of species of insects.\cite{Wdisc_nem}

To give a bit of background,\textit{Wolbachia} was first discovered by Samuel Burt Wolbach and described by Marshall Hertig in 1924.\cite{Winit} Wolbach had been studying the cause of Rocky Mountain Spotted Fever and by 1916 had shown that \textit{Dermacentor andersoni}, a tick native to the Rocky Mountain area, was the transmitter of the disease.\cite{wolbachia} He was unable to grow the actual bacteria causing the disease, a species of \textit{Rickettsia}, in a cell-free culture leading ``him to speculate as to the relationship between the cells of the host and the intracellular parasites."\cite{wolbachia} This is important to note, because the discovery of \textit{Wolbachia} is closely linked to this idea.

@@

Also during the 1960s and '70s scientists noted ``unusual structures in the oocytes or in the hypodermis" of filarial nematodes but did not associate them with bacteria.\cite{wolbachia} It wasn't until 1995 that these ``unusual structures" were determined to be \textit{Wolbachia}.\cite{Wstruct}

@@

%##### How does it function?
The most prominent of these reproductive changes is cytoplasmic incompatibility which is a ``reproductive incompatibility between sperm and egg."\cite{Wbio} \textit{Wolbachia} can only be passed on through eggs, and not sperm, so in order to get around this limitation it has evolved to not allow inter-breeding between members of the same species where only one of them is infected with \textit{Wolbachia}. However, it gets even more complicated than this.

@@

The exact mechanisms for the cause of cytoplasmic incompatibility are not known. What is known is that the \textit{Wolbachia} in the male modify the sperm in such a way that only if the same strain of \textit{Wolbachia} is in the female species to ``recover" the sperm will correct fertilization occur.\cite{Wbio}

In many ways \textit{Wolbachia} acts like an encryption algorithm, ``encrypting" the sperm sent to the female, and only if the female has the correct decryption algorithm can she get pregnant. This is also why different strains of \textit{Wolbachia} are incompatible, using a wrong decryption algorithm will generate nonsense.

@@

Perhaps the most famous case of parthogenisis caused by \textit{Wolbachia} is in the \textit{Trichogramma} wasps, where all members of the species are female. Weirdly, by treating the wasps with antibiotics to kill of the \textit{Wolbachia} cause ``some female parthenogenetic strains ... to revert to production of male progeny."\cite{Wpar_removal} It would appear that \textit{Wolbachia} is merely surpressing the need for male's in certain species. Even stranger is the fact that ``phylogenetic evidence suggests that [parthogenisis] has evolved several times independently in these bacteria" which, as one scientist puts it, suggests ``a simple biochemical mechanism."\cite{Wbio}

@@

\section*{Toxoplasma}
\textit{Toxoplasma gondii} is a “coccidian parasite of cats with all non-feline warm blooded animals (including humans) as intermediate hosts” \cite{Tlife_cycle}. It was first discovered by scientists in 1908 in the rodent Ctenodactylus gundi, known commonly as the North African Gundi \cite{Tlife_cycle}. In 1917, it was discovered that gundis were not infected naturally, but through life in captivity. This raised the question of transmission, and transmission remained mysterious for years, as scientists tried again and again to understand it but came up with “essentially unsuccessful results” \cite{Tlife_cycle}. Now, thanks to the work of several scientists between the 1960s and late 1990s, we understand that \textit{T. gondii} is generally contracted by the ingestion of infected meat or through contact with cat faeces \cite{Tlife_cycle}. We also have begun looking into transmission via birds, which was not initially considered because all other hosts thus far discovered have been mammals \cite{Tbirds}, and the mechanics behind how dogs might transfer the bacteria from cat faeces to humans \cite{Tdogs}.

We also understand now that \textit{T. gondii} infects and reproduces sexually within cats, but all other mammals can also carry the bacteria and become affected by it. Once \textit{Toxoplasma gondii} has entered the body, it makes its way to the lining of the intestines. There, it begins to replicate. In non-feline mammals such as mice, the bacteria reproduces asexually and begins to move throughout the body. Migration into the bloodstream allows the bacteria to reach all parts of the body, including soft tissue, the lymph nodes, and the brain. Tissue cysts begin to form and \textit{T. gondii} continues to replicate \cite{Tmice}. In cats, \textit{T. gondii} reproduces sexually and its oocysts are shed by the cat in its feces. Transmission of these oocysts from the feces to other mammals, either via direct contact or some proxy, is what allows the bacteria to infect its intermediate hosts. Strangely enough, cats have shown signs of immunity to this shedding. Cats do not appear to be negatively affected by \textit{T. gondii's} presence, however there is evidence that a cat which has shed oocysts once has an immunity to shedding them again \cite{Tcat_immunity}. A study was done to test whether or not this immunity lasts for 6 years, and it in fact appears to, unless the cat comes into contact with a more virulent strain of \textit{T. gondii} \cite{Tcat_immunity}. This is an intriguing phenomenon – how can \textit{T. gondii} be effectively spread if cats no longer shed its oocysts after having been infected just once? And why do cats develop this immunity if they are not negatively affected by the bacteria?

Regardless, \textit{Toxoplasma gondii} is spread and its cycle does appear to be effective. \textit{T. gondii's} most common modus operandi is to infect rats and make the rats a tool for re-infecting cats. Rats that are infected by \textit{T. gondii} have been studied and observed to be “significantly more active than their uninfected counterparts,” while rats infected with other parasites that also have life cycles similar to that of \textit{T. gondii} did not show any altered behavior \cite{Trat}. Although “the mechanism of action by which \textit{T. gondii} alters rodent behavior is unknown” \cite{Trat}, this does show that rodent behavior is affected by \textit{Toxoplasma gondii}. The current understanding is that infected rats become more active (and therefore more available to cats), while simultaneously losing their aversion to the smells associated with cats \cite{Tpodcast_replace}. Rats then become easy targets, so cats eat them and the tissue cysts formed by \textit{T. gondii} travel to the cat’s intestines, where the cycle begins again.

@@

\textit{Ophiocordyceps unilateralis} was first discovered in 1859 by the British naturalist Alfred Russel Wallace who, probably not coincidently, is best known for independently conceiving the theory of evolution through natural selection. Darwin published his now famous ``On the Origin of Species" in the same year, which gives some good context as to where the scientific paradigms were at the time.\cite{darwin}

\section*{Impacts And Analysis}

@@

\textit{Toxoplasma gondii} is a microbe that we may have only discovered 107 years ago, and most people have never heard of it, but it still has a large impact on the human population. ``Infection with \textit{Toxoplasma gondii} can cause severe illness when the organism is contracted congenitally or when it is reactivated in immune-suppressed persons” \cite{Tsero}, which is how it most greatly affects people. Many people afflicted by AIDS or cancer have developed toxoplasmosis due to the weakened state of their immune systems. Children are also at risk for serious complications caused by \textit{T. gondii}, as exemplified by the fact that when research into this microbe began, one of the first examined victims was an infant girl who died at one month old due to congenital toxoplasmosis \cite{Tlife_cycle}.

Humans infected by \textit{T. gondii} can exhibit a variety of symptoms. Most people are not at all aware of their infection. Some experience flu-like symptoms, including swollen lymph nodes and muscle aches. However, serious cases can lead to infections in the eyes, brain, and other organs. Some infants infected congenitally are born with serious eye or brain damage, and others who may not have symptoms right away may develop them later in life \cite{Tsymptoms}. In addition, there is research suggesting that people who own cats during childhood may be more likely to be diagnosed with schizophrenia later in life \cite{Tschiz}. The full impacts of \textit{Toxoplasma gondii} on humans have yet to be quantified and understood, but the impacts of which we are currently aware are substantial.

Let’s not forget that \textit{Toxoplasma gondii} is found everywhere. Birds carry it and can shed the oocysts, spreading the microbe \cite{Tbirds}. There is speculation that dogs might carry it on their fur and aid in the transmission of the microbe to humans \cite{Tdogs}. And scientists have come to believe that ``some 25 percent of the global human population is chronically infected with T. gondii” \cite{TUSDA}. Currently, there is no vaccine that is safe for humans to use against \textit{T. gondii} \cite{Tvaccine}, but research is continuing, particularly at the Toxoplasmosis Research Institute and Center in Chicago, IL. This microbe may have bigger plans for us than we can even imagine at this point, so research into it should continue until we understand it better.

The presence of \textit{Wolbachia} in filarial nematodes has raised interesting questions about how \textit{Wolbachia} evolves in its host species. Some scientists discovered that \textit{Wolbachia} ``have possibly been lost during evolution along some lineages of filarial nematodes."\cite{Wevolution_loss} The reason this discovery is bizare is because \textit{Wolbachia}'s ability to evolve into the life-cycles of so many different species makes sense if \textit{Wolbachia} gives the infected a fitness advantage over uninfected members. However, if species evolve to get rid of \textit{Wolbachia} then this points to the fact that \textit{Wolbachia} does not supply them with a fitness advantage.

@@

Using \textit{Wolbachia} as an ``antidote" for different diseases is novel and has potential to change the world dramatically. The best part is that it is a one-time use antidote, because \textit{Wolbachia} has the tools to infect an entire species through vertical transmission. It may also cause scientists to look at other symbiotic relationships between organisms, and see if they have any potential to help humans. There might be a new era of using living organisms as tools to cure diseases and help humans in other ways.

@@

As great as the three microbes mentioned above are, they are not the ideal candidate for addition to the Six Microbes Class Syllabus. Scientifically there is still so much more to be learned about them, and there is really not much that can be done in a laboratory setting that involves these microbes and is safe or legal. These microbes have been around for a long time, but only recently has research been done on them. They just have not been around long enough to have had an impact yet.

These microbes are incredibly unique, and have awesome characteristics in terms of science. Mind-control and reproductive system hijacking not only sounds awesome, it is awesome. However, the actual details of how these microbes do what they do is still unknown. The general idea is there, but the in-depth analysis of the exact chemical process is not. For now, these microbes have \textit{awesome} characteristics, but there is so much more to learn about them that the science is a little lacking.

Although it would be really enjoyable to watch \textit{Cordyceps} fungii growing out of an ant’s skull, it would not make the most active laboratory study. \textit{Wolbachia} may not be the best idea because bringing a bunch of wasps into a small, enclosed lab room has a lot of potential for things to go wrong. Likewise, infecting students with \textit{Toxoplasma} to test if they develop schizophrenia would be interesting -- although possibly not legal. In general, these microbes do not lend themselves easily towards laboratory studies, especially ones that are meant to be conducted over one or two weeks.

The oldest evidence that we have of the \textit{Cordyceps} fungus influencing the mind of ants are some fossilized leaves with the distinct marks of the signature “death grip” on them. Radiometric dating of the fossils shows that they are approximately 47 millions years old, making them the oldest known record of behavioral manipulation.\cite{old_leaf}

Despite having evolved behavioral manipulation abilities so long ago, the fungi have not particularly expanded their impact on the world. 47 million years ago the closest animal to a human on earth was a group of dry-nosed monkeys.

Like the \textit{Cordyceps}, the \textit{Wolbachia} and \textit{Toxoplasma} have been around for a while, although scientists do not know for sure if they have changed all that much. In general, all three microbes have not had much of a historical impact on humans. There are no mind controlled humans that can be linked to either \textit{Toxoplasma} or \textit{Cordyceps}, and as far as we know there is no \textit{Wolbachia} living inside of us. The impacts have yet to be historical for these microbes, and in fact they only exist in the future.

    Although these microbes have had a massive impact on how life has developed, they have not yet had a massive impact on human life (at least that we can describe). However, these microbes are really only just beginning to be researched, and the potential they bring for changing the world is very high. From curing diseases and making medicine to helping us understand evolution, these microbes \textit{will} change the world. We just have not been studying them long enough to know \textit{how} just yet.
